Randomized trace estimation for parameter-dependent matrices applied to spectral density approximation

### 2024-07-09 paper/analysis.tex
pull changes from overleaf

Just as in the constant matrix case, our goal is now to prove that the Nyström++ estimator \refequ{equ:nystrompp-trace-estimator} achieves an $\varepsilon$-relative error with just $\mathcal{O}(\varepsilon^{-1})$ random vectors used to construct the estimate. To do so, we mimic the proof of a similar bound for the Hutch++ estimator \cite[Theorem 3.1]{meyer-2021-hutch-optimal}. Correspondingly, we will need to show that with $n_{\mtx{\Omega}} = \mathcal{O}(\varepsilon^{-2})$ random vectors, the Nyström approximation verifies $\int \lVert \mtx{B}(t) - \Nystr{\mtx{\Omega}}{\mtx{B}}(t)\rVert _F~\mathrm{d}t \leq \varepsilon \int \Trace(\mtx{B}(t))~\mathrm{d}t$, regardless of the structure of the matrix $\mtx{B}(t)$. Just as in the corresponding proof for the Girard-Hutchinson estimator \refthm{thm:hutchinson}, we again bound the higher order moments of the approximation error for fixed $t$ and then use Minkowsi's integral inequality to carry over the results to the continuous case. We will do this in a sequence of three lemmas.

@@

\begin{proof}
    In the proof of \cite[Lemma B.1]{tropp-2023-randomized-algorithms}, it is shown that Slepian's inequality \cite[Theorem 7.2.1]{vershynin-2018-highdimensional-probability} applied to two appropriately constructed Gaussian random fields yields
    \begin{equation}
        \mathbb{E}^{\sfrac{p}{2}}\left[ \lVert \mtx{A} \mtx{\Omega} \rVert _2^2 \right]
        \leq \lVert \mtx{A} \rVert _2^2 \mathbb{E}^{\sfrac{p}{2}}\left[ \lVert \vct{\omega}_1 \rVert _2^2 \right] + \mathbb{E}^{\sfrac{p}{2}}\left[ \lVert \mtx{A} \vct{\omega}_2 \rVert _2^2 \right]
    \end{equation}
    for independent standard Gaussian random vectors $\vct{\omega}_1 \in \mathbb{R}^{k}$ and $\vct{\omega}_2 \in \mathbb{R}^{m}$. Both moments can be bound using \reflem{lem:spectral-norm-moment-vector}, one of which with $\mtx{A} = \mtx{I}_k$, to get
    \begin{align}
        \mathbb{E}^{\sfrac{p}{2}}\left[ \lVert \mtx{A} \mtx{\Omega} \rVert _2^2 \right]
        &\leq \lVert \mtx{A} \rVert _2^2 (k + p) \left( \lVert \mtx{I}_k \rVert _2^2 + \frac{1}{k} \lVert \mtx{I}_k \rVert _F^2 \right) + (k + p) \left( \lVert \mtx{A} \rVert _2^2 + \frac{1}{k} \lVert \mtx{A} \rVert _F^2 \right) \notag \\
        &\leq (k + p) \left( 3 \lVert \mtx{A} \rVert _2^2 + \frac{1}{k} \lVert \mtx{A} \rVert _F^2 \right).
    \end{align}

@@

\begin{lemma}{Nyström approximation for parameter-dependent matrices}{nystrom}
    Let $\mtx{B}(t) \in \mathbb{R}^{n \times n}$ be symmetric positive semi-definite and continuously depend on $t \in [a, b]$. Then its Nyström approximation $\Nystr{\mtx{\Omega}}{\mtx{B}}(t)$ with Gaussian random matrix $\mtx{\Omega} \in \mathbb{R}^{n \times n_{\mtx{\Omega}}}$ for \emph{even} $n_{\mtx{\Omega}} \geq 4$ verifies with probability at least $1 - \gamma^{-\sfrac{n_{\mtx{\Omega}}}{4}}$
    \begin{equation}
        \int_{a}^{b} \lVert \mtx{B}(t) - \Nystr{\mtx{\Omega}}{\mtx{B}}(t) \rVert _F~\mathrm{d}t \leq \gamma \frac{c}{\sqrt{n_{\mtx{\Omega}}}} \int_{a}^{b} \Trace(\mtx{B}(t))~\mathrm{d}t
        \label{qu:nystrompp-theorem-bound}
    \end{equation}
    for any $\gamma \geq 1$ and a universal constant $c > 0$. In particular, if we choose $n_{\mtx{\Omega}} = \mathcal{O}(\varepsilon^{-2} + \log(\delta^{-1}))$, then with probability at least $1-\delta$ for $\delta \in (0, 1)$ and $\varepsilon > 0$ we have $\int_{a}^{b} \lVert \mtx{B}(t) - \Nystr{\mtx{\Omega}}{\mtx{B}}(t) \rVert _F~\mathrm{d}t \leq \varepsilon \int_{a}^{b} \Trace(\mtx{B}(t))~\mathrm{d}t$.
\end{lemma}

>>>>

\refthm{thm:hutchinson} has established a relative $\varepsilon$-error for the Girard-Hutchinson estimator with $\mathcal{O}(\varepsilon^{-2})$ queries. The aim of this section is to show that the Nyström++ estimator~\refequ{equ:nystrompp-trace-estimator} improves this to $\mathcal{O}(\varepsilon^{-1})$ matrix-vector products with random vectors, \emph{without} requiring any assumption on the singular values of $\mtx{B}(t)$. This parallels existing results for the constant case
for Hutch++\cite[Theorem 3.1]{meyer-2021-hutch-optimal} and Nyström++\textcolor{red}{??? Include specific result ???}.

To do so, we mimic the proof of a similar bound for the Hutch++ estimator . Correspondingly, we will need to show that with $n_{\mtx{\Omega}} = \mathcal{O}(\varepsilon^{-2})$ random vectors, the Nyström approximation verifies $\int \lVert \mtx{B}(t) - \Nystr{\mtx{\Omega}}{\mtx{B}}(t)\rVert _F~\mathrm{d}t \leq \varepsilon \int \Trace(\mtx{B}(t))~\mathrm{d}t$, regardless of the structure of the matrix $\mtx{B}(t)$. Just as in the corresponding proof for the Girard-Hutchinson estimator \refthm{thm:hutchinson}, we again bound the higher order moments of the approximation error for fixed $t$ and then use Minkowsi's integral inequality to carry over the results to the continuous case. We will do this in a sequence of three lemmas.

@@

\begin{proof}
    In the proof of \cite[Lemma B.1]{tropp-2023-randomized-algorithms}, it is shown that Slepian's inequality \cite[Theorem 7.2.1]{vershynin-2018-highdimensional-probability} applied to two appropriately constructed Gaussian random fields yields
    \begin{equation}
        \mathbb{E}^{\sfrac{p}{2}}\left[ \lVert \mtx{A} \mtx{\Omega} \rVert _2^2 \right]
        \leq \lVert \mtx{A} \rVert _2^2 \mathbb{E}^{\sfrac{p}{2}}\left[ \lVert \vct{\omega}_1 \rVert _2^2 \right] + \mathbb{E}^{\sfrac{p}{2}}\left[ \lVert \mtx{A} \vct{\omega}_2 \rVert _2^2 \right]
    \end{equation}
    for independent standard Gaussian random vectors $\vct{\omega}_1 \in \mathbb{R}^{k}$ and $\vct{\omega}_2 \in \mathbb{R}^{m}$. \textcolor{red}{Doesn't A.1 give a tighter bound for the first term?} Both moments can be bound using \reflem{lem:spectral-norm-moment-vector}, one of which with $\mtx{A} = \mtx{I}_k$, to get
    \begin{align}
        \mathbb{E}^{\sfrac{p}{2}}\left[ \lVert \mtx{A} \mtx{\Omega} \rVert _2^2 \right]
        &\leq \lVert \mtx{A} \rVert _2^2 (k + p) \left( \lVert \mtx{I}_k \rVert _2^2 + \frac{1}{k} \lVert \mtx{I}_k \rVert _F^2 \right) + (k + p) \left( \lVert \mtx{A} \rVert _2^2 + \frac{1}{k} \lVert \mtx{A} \rVert _F^2 \right) \notag \\
        &\leq (k + p) \left( 3 \lVert \mtx{A} \rVert _2^2 + \frac{1}{k} \lVert \mtx{A} \rVert _F^2 \right).
    \end{align}

@@

\color{blue}

\begin{lemma}{Nyström approximation for parameter-dependent matrices}{nystrom}
    Let $\mtx{B}(t) \in \mathbb{R}^{n \times n}$ be symmetric positive semi-definite and continuously depend on $t \in [a, b]$. Consider the Nyström approximation $\Nystr{\mtx{\Omega}}{\mtx{B}}(t)$ with a Gaussian random matrix $\mtx{\Omega} \in \mathbb{R}^{n \times n_{\mtx{\Omega}}}$ and \emph{even} $n_{\mtx{\Omega}} \geq 4$.
    Then for any $\gamma \geq 1$ the bound
    \begin{equation}
        \int_{a}^{b} \lVert \mtx{B}(t) - \Nystr{\mtx{\Omega}}{\mtx{B}}(t) \rVert _F~\mathrm{d}t \leq \gamma \frac{c}{\sqrt{n_{\mtx{\Omega}}}} \int_{a}^{b} \Trace(\mtx{B}(t))~\mathrm{d}t
        \label{qu:nystrompp-theorem-bound}
    \end{equation}
        holds with probability at least $1 - \gamma^{-\sfrac{n_{\mtx{\Omega}}}{4}}$ for 
        a universal constant $c > 0$. In particular, given $\varepsilon > 0$ and $\delta \in (0, 1)$, the bound $\int_{a}^{b} \lVert \mtx{B}(t) - \Nystr{\mtx{\Omega}}{\mtx{B}}(t) \rVert _F~\mathrm{d}t \leq \varepsilon \int_{a}^{b} \Trace(\mtx{B}(t))~\mathrm{d}t$ holds with probability at least $1-\delta$ if $n_{\mtx{\Omega}} = \mathcal{O}(\varepsilon^{-2} + \log(\delta^{-1}))$.
\end{lemma}
